### 2017-06-01 paper/prelims.tex
added a definition of LexRSM

\vspace{-1em}
\subsection{Almost-Sure and Probabilistic Termination}
\vspace{-0.5em}
We consider computational problems related to the basic liveness properties of 
\APP{s}, namely \emph{termination} and its generalization, \emph{reachability}.

\smallskip\noindent{\em Termination, reachability, and termination time.}
In the following, consider an \APP{} $P$ and its associated pCFG $\pCFG_P$. 
We say that a run $\run$ of $\pCFG_P$ \emph{reaches} a set of configurations 
$C$ if it  contains a configuration from $C$. A run \emph{terminates} if it reaches a 
configuration whose first component (i.e. a location of $\pCFG_P$) is the 
location
$\loc_{P}^{\lout}$ corresponding to the value of the program counter after 
executing $P$. 
To each set of configurations $C$ we can assign a random variable $\treach{C}$ 
such that for each run $\run$ the value $\treach{C}(\run)$ represents the first 
point in time when 
the current configuration on $\run$ is in $C$. If a run $\run$ does \emph{not} 
reach a set $C$, then 
$\treach{C}(\run)=\infty$. We call $\treach{C}$ the 
\emph{reachability time} of $C$. In particular, if $C$ is the set 
of all configurations $(\loc,\vec{x})$ such that $\loc=\loc^\lout_P$ (the 
terminal location of $\pCFG_P$), then $\treach{C}$ is 
called a \emph{termination time}, as it returns the number of steps after which 
$\run$ terminates. Since termination time is an important concept on its own, 
we use a special notation $\ttime$ for it.
%\begin{example}
%Consider the program $P$ such that \hspace{-0.25cm}\newsavebox{\figa}
%\lstset{language=affprob}
%\lstset{tabsize=1}
%\begin{lrbox}{\figa}
%\begin{lstlisting}[mathescape]
%$P=$x:=2;y:=1
%\end{lstlisting}
%\end{lrbox}
%\usebox{\figa}. The associated pCFG has three locations: initial location 
%$\loc_{P}^{\lin}$, terminal location $\loc_{P}^{\lout}$, and one 
%``intermediate'' location, say $\loc$. In the pCFG consider a run 
%$\run=(\loc_{P}^{\lin},(0,0))(\loc,(2,0))(\loc_{P}^{\lout},(2,1))(\loc_{P}^{\lout},(2,1))\cdots$.
% Then $\ttime(\run)=2$. If we denote by $C$ the set of configurations 
% $\{(\loc',(x,y))\mid \loc' \in\{\loc_{P}^{\lin},\loc_{P}^{\lout},\loc\} \text{ 
% and } x\geq 2\}$, then $\treach{C}(\run)=1$.
%\end{example}
Since a probabilistic program may exhibit more than one run, we are interested 
in probabilities of runs that terminate or reach some set of configurations. 
This gives rise to the following fundamental computational problems regarding 
termination:

\begin{compactenum}
\item \emph{Almost-sure termination:} A probabilistic program $P$ is 
almost-surely (a.s.) 
terminating if under each scheduler $\sigma$ it holds that 
$\probm^{\sigma}(\{\run\mid \run \text{ terminates}\}) = 1$, or equivalently, 
if for each $\sigma$ it holds $\probm^{\sigma}(\ttime<\infty)=1$. In 
almost-sure termination question for $P$ we aim to prove that $P$ 
is 
almost-surely terminating.
\item \emph{Probabilistic termination:} In probabilistic termination question 
for $P$ we aim to compute a \emph{lower bound on the probability of 
termination}, i.e. a bound $b\in[0,1]$ such that for each scheduler 
$\sigma$ it holds $\probm^{\sigma}(\{\run\mid \run \text{ terminates}\}) \geq b$ 
(or equivalently $\probm^{\sigma}(\ttime<\infty)\geq b$).
\end{compactenum}

@@

\begin{compactenum}
\item \emph{Almost-sure reachability:} For a set $C$ of configurations of a
probabilistic program $P$, prove (if possible) that under each scheduler 
$\sigma$ it holds that
$\probm^{\sigma}(\treach{C}<\infty)=1$.
\item \emph{Probabilistic reachability:} For a set $C$ of configurations of 
a probabilistic program $P$, compute a bound $b\in[0,1]$ such that for each scheduler
$\sigma$ it holds $\probm^{\sigma}(\treach{C}<\infty)\geq b$.
\end{compactenum}

>>>>

\vspace{-1em}
\subsection{Almost-Sure and Positive Termination}
\vspace{-0.5em}
We consider computational problems related to the basic liveness properties of 
\APP{s}, namely \emph{termination} and its generalization, \emph{reachability}.

\smallskip\noindent{\em Termination, reachability, and termination time.}
In the following, consider an \APP{} $P$ and its associated pCFG $\pCFG_P$. 
We say that a run $\run$ of $\pCFG_P$ \emph{reaches} a set of configurations 
$\confset$ if it  contains a configuration from $\confset$. A run 
\emph{terminates} if it reaches a 
configuration whose first component (i.e. a location of $\pCFG_P$) is the 
location
$\loc_{P}^{\lout}$ corresponding to the value of the program counter after 
executing $P$. 
To each set of configurations $\confset$ we can assign a random variable 
$\treach{\confset}$ 
such that for each run $\run$ the value $\treach{\confset}(\run)$ represents 
the first 
point in time when 
the current configuration on $\run$ is in $\confset$. If a run $\run$ does 
\emph{not} 
reach a set $\confset$, then 
$\treach{\confset}(\run)=\infty$. We call $\treach{\confset}$ the 
\emph{reachability time} of $\confset$. In particular, if $\confset$ is the set 
of all configurations $(\loc,\vec{x})$ such that $\loc=\loc^\lout_P$ (the 
terminal location of $\pCFG_P$), then $\treach{\confset}$ is 
called a \emph{termination time}, as it returns the number of steps after which 
$\run$ terminates. Since termination time is an important concept on its own, 
we use a special notation $\ttime$ for it.
%\begin{example}
%Consider the program $P$ such that \hspace{-0.25cm}\newsavebox{\figa}
%\lstset{language=affprob}
%\lstset{tabsize=1}
%\begin{lrbox}{\figa}
%\begin{lstlisting}[mathescape]
%$P=$x:=2;y:=1
%\end{lstlisting}
%\end{lrbox}
%\usebox{\figa}. The associated pCFG has three locations: initial location 
%$\loc_{P}^{\lin}$, terminal location $\loc_{P}^{\lout}$, and one 
%``intermediate'' location, say $\loc$. In the pCFG consider a run 
%$\run=(\loc_{P}^{\lin},(0,0))(\loc,(2,0))(\loc_{P}^{\lout},(2,1))(\loc_{P}^{\lout},(2,1))\cdots$.
% Then $\ttime(\run)=2$. If we denote by $C$ the set of configurations 
% $\{(\loc',(x,y))\mid \loc' \in\{\loc_{P}^{\lin},\loc_{P}^{\lout},\loc\} \text{ 
% and } x\geq 2\}$, then $\treach{C}(\run)=1$.
%\end{example}
Since a probabilistic program may exhibit more than one run, we are interested 
in probabilities of runs that terminate or reach some set of configurations. 
This gives rise to the following fundamental computational problems regarding 
termination:

\begin{compactenum}
\item \emph{Almost-sure termination:} A probabilistic program $P$ is 
almost-surely (a.s.) 
terminating if under each scheduler $\sigma$ it holds that 
$\probm^{\sigma}(\{\run\mid \run \text{ terminates}\}) = 1$, or equivalently, 
if for each $\sigma$ it holds $\probm^{\sigma}(\ttime<\infty)=1$. In 
almost-sure termination question for $P$ we aim to prove that $P$ 
is 
almost-surely terminating.
\item \emph{Positive termination:} A probabilistic program $P$ is 
positively
terminating if under each scheduler $\sigma$ it holds that 
$\E^{\sigma}(\{\ttime\}) \infty$. In 
positive termination question for $P$ we aim to prove that $P$ 
is 
positively terminating. Note that each positively terminating program is also 
a.s. terminating, but the converse does not hold.
\end{compactenum}

@@

\begin{compactenum}
\item \emph{Almost-sure reachability:} For a set $C$ of configurations of a
probabilistic program $P$, prove (if possible) that under each scheduler 
$\sigma$ it holds that
$\probm^{\sigma}(\treach{C}<\infty)=1$.
\item \emph{Positive reachability:} For a set $C$ of configurations of 
a probabilistic program $P$, prove (if possible) that under each scheduler 
$\sigma$ it holds that
$E^{\sigma}(\treach{C})<\infty$.
\end{compactenum}
